---
title: "Modern ggplot2"
author: "Teun van den Brand"
format: revealjs
---

```{r}
# library(ggplot2)
# theme_set(theme_gray(base_size = 22))
knitr::opts_chunk$set(dev = "ragg_png", dev.args = list(scaling = 2))
```


## Headings

- Column metadata
- Data dictionary

---

### Column metadata

:::: {.columns}

:::{.column}


- 'Pretty labels' implemented as `"label"` attribute in columns.
- Implemented in Hmisc, tinylabels, haven, labelled & sjlabelled

:::

:::{.column}

```{r}
#| echo: true
#| collapse: true
df <- mtcars

df$mpg <- haven::labelled(df$mpg, label = "Miles per gallon")

head(df$mpg)

attr(df$mpg, "label")
```
:::
::::

---

### Column metadata

:::: {.columns}

:::{.column}

- 'Pretty labels' implemented as `"label"` attribute in columns.
- Implemented in Hmisc, tinylabels, haven, labelled & sjlabelled
- Careful with label attribute stability

:::

:::{.column}

```{r}
#| echo: true
#| collapse: true
df <- mtcars

attr(df$mpg, "label") <- "Miles per gallon"

head(df$mpg)

vctrs::vec_slice(df$mpg, 1:6)
```

:::

::::

---

### Column metadata

Label attribute automatically detected.

```{r}
#| echo: true
library(ggplot2)

ggplot(df, aes(mpg, disp)) +
  geom_point()
```
---

### Data dictionary

Example dictionary for `mtcars`

```{r}
#| echo: true
dict <- tibble::tribble(
  ~column, ~label,                ~unit,    ~note,
  "mpg",   "Efficiency",          "mi/gal", "Gallons are US gallons",
  "cyl",   "Number of cylinders", "",       "",
  "disp",  "Engine Displacement", "in^3",   "",
  "am",    "Transmission",        "",       "0 = automatic, 1 = manual"
)
dict
```

---

### Data dictionary

```{r}
#| echo: true
#| collapse: true
# Format label as named vector
named_vec <- setNames(dict$label, dict$column)
# or:
named_vec <- dplyr::pull(dict, label, name = column)

named_vec
```

---

### Data dictionary

```{r}
#| echo: true
#| code-line-numbers: "3"
ggplot(mtcars, aes(mpg, disp, colour = cyl)) +
  geom_point() +
  labs(dictionary = named_vec)
```

---

### Pros

- Label variables directly, rather than aesthetics
- Rewards habit of annotating data
- Reusable within document

```{r}
#| echo: true
ggplot(mtcars, aes(cyl, mpg, group = cyl)) +
  geom_boxplot() +
  labs(dictionary = named_vec)
```
---

### Cons

- Extra effort for 'naked' data
- Expressions like `factor(cyl)` or `cyl + 1` do not get automatic labels

## Patterns and gradients

In R 4.1 the grid package introduced patterns and gradients.

* `grid::linearGradient()`
* `grid::radialGradient()`
* `grid::pattern()`

We allow these as `fill` aesthetic in ggplot2.

---

### Gradients

Simple examples of linear and radial gradients.

```{r}
#| echo: true
colours <- hcl.colors(100, "Sunset")

library(grid)
linear <- linearGradient(colours, x1 = 0.5, x2 = 0.5)
radial <- radialGradient(rev(colours), cx1 = 0.8, cy1 = 0.8, group = FALSE)
```

---

### Gradients

Use these gradients by providing them as a list.

```{r}
#| echo: true
#| code-line-numbers: "3"
ggplot(mtcars) +
  aes(factor(vs)) +
  geom_bar(fill = list(linear, radial))
```

---

### Patterns

Patterns are more complicated.
You may need to know a little bit of grid to get these right.
Here we're using a diagonal line as a pattern.

```{r}
#| echo: true
#| code-line-numbers: "3-6,9"
width <- height <- unit(3, "mm")

inner_drawing <- segmentsGrob(
  gp = gpar(col = "black"),
  vp = viewport(width = width, height = height)
)

hatching <- pattern(
  inner_drawing,
  width = width, height = height,
  extend = "repeat"
)
```

---

### Patterns

Like gradients, patterns can be given as a list.

```{r}
#| echo: true
#| code-line-numbers: "4"
ggplot(mtcars) +
  aes(factor(vs)) +
  geom_bar(
    fill = list(hatching), 
    colour = "black"
  )
```

---

### Patterns

To 'fix' patterning artefacts, you may need to adjust the strokes in the inner drawing.

```{r}
#| echo: true
#| code-line-numbers: "4-5"
width <- height <- unit(3, "mm")
inner_drawing <- segmentsGrob(
  x0 = c(-1, -1, 0), x1 = c(1, 2, 2),
  y0 = c(0, -1, -1), y1 = c(2, 2, 1),
  gp = gpar(col = "black"),
  vp = viewport(width = width, height = height)
)
hatching <- pattern(
  inner_drawing,
  width = width, height = height,
  extend = "repeat"
)
```


```{r}
grid.newpage()
grid.draw(inner_drawing)
grid.rect(
  width = width, height = height, 
  gp = gpar(fill = NA, lty = "dotted", col = "grey50")
)
```

---

### Patterns

Using patterns as a scale.

```{r}
#| echo: true
#| code-line-numbers: "3-5"
ggplot(mtcars, aes(cyl, disp, fill = factor(cyl))) +
  geom_boxplot() +
  scale_fill_manual(
    values = list(linear, radial, hatching)
  )
```

---

### Ribbon gradient

Ribbon geometries now render a varying `fill` aesthetic as a gradient.

```{r}
#| echo: true
ggplot(economics) +
  aes(date, unemploy, fill = uempmed) +
  geom_area()
```

---

### Patterns galore

Using the gridpattern package to easily generate patterns.

```{r}
#| echo: true
herringbone <- gridpattern::patternFill(
  "polygon_tiling", type = "herringbone", 
  spacing = 0.2, units = "cm", colour = "grey40", linewidth = 0.3
)
hexagons <- gridpattern::patternFill(
  "polygon_tiling", type = "hexagonal",
  spacing = 0.2, units = "cm", colour = "grey40", linewidth = 0.3
)
waves <- gridpattern::patternFill(
  "wave", spacing = 0.2, units = "cm", colour = "grey40", linewidth = 0.3
)
```

---

### Patterns galore

Using the gridpattern package to easily generate patterns.

```{r}
#| echo: true
#| code-fold: true
ggplot(mtcars, aes(cyl, disp, fill = factor(cyl))) +
  geom_boxplot() +
  scale_fill_manual(
    values = list(herringbone, hexagons, waves)
  )
```

### Patterns galore

Parametrised patterns with the ggpattern package.

```{r}
#| echo: true
library(ggpattern)

ggplot(mtcars, aes(cyl, disp, pattern_spacing = cyl, group = cyl)) +
  geom_boxplot_pattern(pattern_fill = "black") +
  scale_pattern_spacing_continuous(range = c(0.01, 0.05))
```

---

### Patterns galore

Emoji isotype plot using text patterns.

```{r}
#| echo: true
#| code-fold: true
# Helper function
width <- unit(20, "pt")
patternise_text <- function(text) {
  lapply(text, function(string) {
    grob <- textGrob(string, x = 0, hjust = 0, gp = gpar(fontsize = 18))
    pattern(
      grob,
      x = 0, hjust = 0,
      width = width, 
      extend = "repeat", 
      # Center text per bar using height/group
      height = unit(1, "npc"),
      group = FALSE
    )
  })
}

# Stats for the Netherlands
df <- data.frame(
  animal = c("chickens", "pigs",   "cows", "sheep", "goats", "humans"),
  amount = c(99900000,   11400000, 3800000, 850000, 480000,  17990000)
)

ggplot(df, aes(amount, animal, fill = animal)) +
  geom_col() +
  scale_fill_manual(
    values = patternise_text(c(
      "chickens" = "ðŸ“",
      "pigs"     = "ðŸ–",
      "cows"     = "ðŸ„",
      "sheep"    = "ðŸ‘",
      "goats"    = "ðŸ",
      "humans"   = "ðŸ§"
    ))
  ) +
  scale_x_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "M")
  ) +
  theme(
    legend.key.width = width,
    legend.key.height = unit(18, "pt") # see fontsize in pattern
  )
```

