---
title: "Modern ggplot2"
author: "Teun van den Brand"
format: revealjs
---

```{r}
# library(ggplot2)
# theme_set(theme_gray(base_size = 22))
knitr::opts_chunk$set(dev = "ragg_png", dev.args = list(scaling = 2))
```


## Headings

- Column metadata
- Data dictionary

---

### Column metadata

:::: {.columns}

:::{.column}


- 'Pretty labels' implemented as `"label"` attribute in columns.
- Implemented in Hmisc, tinylabels, haven, labelled & sjlabelled

:::

:::{.column}

```{r}
#| echo: true
#| collapse: true
df <- mtcars

df$mpg <- haven::labelled(df$mpg, label = "Miles per gallon")

head(df$mpg)

attr(df$mpg, "label")
```
:::
::::

---

### Column metadata

:::: {.columns}

:::{.column}

- 'Pretty labels' implemented as `"label"` attribute in columns.
- Implemented in Hmisc, tinylabels, haven, labelled & sjlabelled
- Careful with label attribute stability

:::

:::{.column}

```{r}
#| echo: true
#| collapse: true
df <- mtcars

attr(df$mpg, "label") <- "Miles per gallon"

head(df$mpg)

vctrs::vec_slice(df$mpg, 1:6)
```

:::

::::

---

### Column metadata

Label attribute automatically detected.

```{r}
#| echo: true
library(ggplot2)

ggplot(df, aes(mpg, disp)) +
  geom_point()
```
---

### Data dictionary

Example dictionary for `mtcars`

```{r}
#| echo: true
dict <- tibble::tribble(
  ~column, ~label,                ~unit,    ~note,
  "mpg",   "Efficiency",          "mi/gal", "Gallons are US gallons",
  "cyl",   "Number of cylinders", "",       "",
  "disp",  "Engine Displacement", "in^3",   "",
  "am",    "Transmission",        "",       "0 = automatic, 1 = manual"
)
dict
```

---

### Data dictionary

```{r}
#| echo: true
#| collapse: true
# Format label as named vector
named_vec <- setNames(dict$label, dict$column)
# or:
named_vec <- dplyr::pull(dict, label, name = column)

named_vec
```

---

### Data dictionary

```{r}
#| echo: true
#| code-line-numbers: "3"
ggplot(mtcars, aes(mpg, disp, colour = cyl)) +
  geom_point() +
  labs(dictionary = named_vec)
```

---

### Pros

- Label variables directly, rather than aesthetics
- Rewards habit of annotating data
- Reusable within document

```{r}
#| echo: true
ggplot(mtcars, aes(cyl, mpg, group = cyl)) +
  geom_boxplot() +
  labs(dictionary = named_vec)
```
---

### Cons

- Extra effort for 'naked' data
- Expressions like `factor(cyl)` or `cyl + 1` do not get automatic labels

## Patterns and gradients

In R 4.1 the grid package introduced patterns and gradients.

* `grid::linearGradient()`
* `grid::radialGradient()`
* `grid::pattern()`

We allow these as `fill` aesthetic in ggplot2.

---

### Gradients

Simple examples of linear and radial gradients.

```{r}
#| echo: true
colours <- hcl.colors(100, "Sunset")

library(grid)
linear <- linearGradient(colours, x1 = 0.5, x2 = 0.5)
radial <- radialGradient(rev(colours), cx1 = 0.8, cy1 = 0.8, group = FALSE)
```

---

### Gradients

Use these gradients by providing them as a list.

```{r}
#| echo: true
#| code-line-numbers: "3"
ggplot(mtcars) +
  aes(factor(vs)) +
  geom_bar(fill = list(linear, radial))
```

---

### Patterns

Patterns are more complicated.
You may need to know a little bit of grid to get these right.
Here we're using a diagonal line as a pattern.

```{r}
#| echo: true
#| code-line-numbers: "3-6,9"
width <- height <- unit(3, "mm")

inner_drawing <- segmentsGrob(
  gp = gpar(col = "black"),
  vp = viewport(width = width, height = height)
)

hatching <- pattern(
  inner_drawing,
  width = width, height = height,
  extend = "repeat"
)
```

---

### Patterns

Like gradients, patterns can be given as a list.

```{r}
#| echo: true
#| code-line-numbers: "4"
ggplot(mtcars) +
  aes(factor(vs)) +
  geom_bar(
    fill = list(hatching), 
    colour = "black"
  )
```

---

### Patterns

To 'fix' patterning artefacts, you may need to adjust the strokes in the inner drawing.

```{r}
#| echo: true
#| code-line-numbers: "4-5"
width <- height <- unit(3, "mm")
inner_drawing <- segmentsGrob(
  x0 = c(-1, -1, 0), x1 = c(1, 2, 2),
  y0 = c(0, -1, -1), y1 = c(2, 2, 1),
  gp = gpar(col = "black"),
  vp = viewport(width = width, height = height)
)
hatching <- pattern(
  inner_drawing,
  width = width, height = height,
  extend = "repeat"
)
```


```{r}
grid.newpage()
grid.draw(inner_drawing)
grid.rect(
  width = width, height = height, 
  gp = gpar(fill = NA, lty = "dotted", col = "grey50")
)
```

---

### Patterns

Using patterns as a scale.

```{r}
#| echo: true
#| code-line-numbers: "3-5"
ggplot(mtcars, aes(cyl, disp, fill = factor(cyl))) +
  geom_boxplot() +
  scale_fill_manual(
    values = list(linear, radial, hatching)
  )
```

---

### Ribbon gradient

Ribbon geometries now render a varying `fill` aesthetic as a gradient.

```{r}
#| echo: true
ggplot(economics) +
  aes(date, unemploy, fill = uempmed) +
  geom_area()
```

---

### Patterns galore

Using the gridpattern package to easily generate patterns.

```{r}
#| echo: true
herringbone <- gridpattern::patternFill(
  "polygon_tiling", type = "herringbone", 
  spacing = 0.2, units = "cm", colour = "grey40", linewidth = 0.3
)
hexagons <- gridpattern::patternFill(
  "polygon_tiling", type = "hexagonal",
  spacing = 0.2, units = "cm", colour = "grey40", linewidth = 0.3
)
waves <- gridpattern::patternFill(
  "wave", spacing = 0.2, units = "cm", colour = "grey40", linewidth = 0.3
)
```

---

### Patterns galore

Using the gridpattern package to easily generate patterns.

```{r}
#| echo: true
#| code-fold: true
ggplot(mtcars, aes(cyl, disp, fill = factor(cyl))) +
  geom_boxplot() +
  scale_fill_manual(
    values = list(herringbone, hexagons, waves)
  )
```

### Patterns galore

Parametrised patterns with the ggpattern package.

```{r}
#| echo: true
library(ggpattern)

ggplot(mtcars, aes(cyl, disp, pattern_spacing = cyl, group = cyl)) +
  geom_boxplot_pattern(pattern_fill = "black") +
  scale_pattern_spacing_continuous(range = c(0.01, 0.05))
```

---

### Patterns galore

Emoji isotype plot using text patterns.

```{r}
#| echo: true
#| code-fold: true
# Helper function
width <- unit(20, "pt")
patternise_text <- function(text) {
  lapply(text, function(string) {
    grob <- textGrob(string, x = 0, hjust = 0, gp = gpar(fontsize = 18))
    pattern(
      grob,
      x = 0, hjust = 0,
      width = width, 
      extend = "repeat", 
      # Center text per bar using height/group
      height = unit(1, "npc"),
      group = FALSE
    )
  })
}

# Stats for the Netherlands
df <- data.frame(
  animal = c("chickens", "pigs",   "cows", "sheep", "goats", "humans"),
  amount = c(99900000,   11400000, 3800000, 850000, 480000,  17990000)
)

ggplot(df, aes(amount, animal, fill = animal)) +
  geom_col() +
  scale_fill_manual(
    values = patternise_text(c(
      "chickens" = "ðŸ“",
      "pigs"     = "ðŸ–",
      "cows"     = "ðŸ„",
      "sheep"    = "ðŸ‘",
      "goats"    = "ðŸ",
      "humans"   = "ðŸ§"
    ))
  ) +
  scale_x_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "M")
  ) +
  theme(
    legend.key.width = width,
    legend.key.height = unit(18, "pt") # see fontsize in pattern
  )
```

## Delayed evaluation

With regards to evaluation, there are three stages:

1. Direct input at start
2. After computing stat
3. After scale mapping

---

### Direct input

Data available from the start, when mapped from data columns.

```{r}
#| echo: true
#| eval: false
aes(x = displ, y = hwy)
```

Unmapped aesthetics like `geom_bar(fill = "red")` are not direct input.  
Data columns that are not mapped are not input.

---

### After computing stat

In addition to aesthetics, *computed variables* become available.  
These are listed in a documentation section in e.g. `?stat_density`.  
Sometimes, these are used by default and can be recognised by the use of `after_stat()`.

```{r}
#| echo: true
StatDensity$default_aes
```

---

### After computing stat

You can use `after_stat()` yourself to redirect or modify stat outputs.

```{r}
#| echo: true
#| code-line-numbers: "4"
ggplot(mpg, aes(displ, drv)) +
  stat_density(
    geom = "tile", position = "identity",
    aes(fill = after_stat(scaled))
  )
```

---

### After computing stat

You may have run into a histogram/density misalignment problem.

```{r}
#| echo: true
ggplot(faithful, aes(waiting)) +
  geom_histogram(binwidth = 2) +
  geom_density()
```

---

### After computing stat

This can be fixed by using the `density` computed variable in the histogram.

```{r}
#| echo: true
#| code-line-numbers: "3"
ggplot(faithful, aes(waiting)) +
  geom_histogram(
    aes(y = after_stat(density)), 
    binwidth = 2
  ) +
  geom_density()
```

---

### After computing stat

Or scaling the `count` computed variable in the density.

```{r}
#| echo: true
#| code-line-numbers: "5"
binwidth <- 2
ggplot(faithful, aes(waiting)) +
  geom_histogram(binwidth = binwidth) +
  geom_density(
    aes(y = after_stat(count * binwidth))
  )
```

---

### After scales

At this stage in the plot, the variables are mapped by the scales.  
It means that the hex codes from colour scales are available, the sizes and shapes of points and other variables mapped by scales.  
You can intervene by using the `after_scale()` function.

--- 

### After scales

A typical use of `after_scale()` is to derive colours from `colour` to `fill` or vice versa.

```{r}
#| echo: true
#| code-line-numbers: "3"
ggplot(mpg, aes(displ, fill = drv)) +
  geom_density(
    aes(colour = after_scale(scales::col_mix(fill, "black")))
  )
```

--- 

### After scales

A nice benefit of using `after_scale()` is that you derive colours, so you can still swap out scales.

```{r}
#| echo: true
last_plot() + 
  scale_fill_viridis_d()
```

### After scales

Another use case can be to create half-geometries.

```{r}
#| echo: true
#| warning: false
ggplot(mpg, aes(class, displ)) +
  geom_boxplot(aes(xmin = after_scale(x)), staplewidth = 0.3) +
  geom_violin(aes(xmax = after_scale(x)))
```

### Staging

When you need a combination of direct input, after stat or after scale modifications, you can use `stage()`.

* `stage(x)` is equivalent to `x`.
* `stage(after_stat = x)` is equivalent to `after_stat(x)`.
* `stage(after_scale = x)` is equivalent to `after_scale(x)`.

### Staging

A typical use case is when you want to initialise the aesthetic with one column, and later modify the mapped values.

```{r}
#| echo: true
ggplot(mpg, aes(class, displ)) +
  geom_violin(
    aes(fill = stage(class, after_scale = scales::col_mix(fill, "transparent")))
  )
```

### Staging

Another use case is to reposition labels after computing a statistic.

```{r}
#| echo: true
#| code-line-numbers: "4-8|9-12"
ggplot(mpg, aes(class, displ, fill = class)) +
  geom_violin() +
  stat_summary(
    fun.data = ~ data.frame(
      mean = mean(.x), 
      sd   = sd(.x), 
      max  = max(.x)
    ),
    aes(
      y = stage(displ, after_stat = max + 0.2),
      label = after_stat(sprintf("%.2fÂ±%.2f", mean, sd))
    ),
    geom = "text"
  )
```

