---
title: "Styling your plot"
author: "Teun van den Brand"
---

```{r}
#| include: false

systemfonts::require_font("Barrio")
systemfonts::require_font("Dancing Script")

library(ggplot2)
theme_set(
  theme_gray() +
    theme(
      panel.background = element_rect(colour = NA, fill = "#F0F1EB")
    )
)
```

## The theme menu

* Ink & Paper
* Default palettes
* Build your own theme
* Custom guides

---

## Ink & Paper

* Complete themes
* Geom element
* Use in layers

---

### Complete themes

```{r}
library(ggplot2)

p <- ggplot(mpg) +
  aes(displ, hwy) +
  geom_point() +
  geom_smooth(
    aes(displ, hwy),
    formula = y ~ x, method = "lm",
    inherit.aes = FALSE
  ) +
  facet_wrap(~ drv) +
  labs(dictionary = c(
      cty = "City\nEfficiency",
      hwy = "Highway Efficiency",
      displ = "Engine Displacement",
      year = "Year",
      "factor(year)" = "Year",
      class = "Class"
  ))
p
```

---

### Complete themes

`ink` affects all foreground elements and parts of layers.

```{r}
#| output-location: default
#| fig-width: 18
p + theme_gray(ink = "red")
```

---

### Complete themes

`paper` affects all background elements and parts of layers.

```{r}
#| output-location: default
#| fig-width: 18
p + theme_bw(paper = "cornsilk")
```

---

### Complete themes

`ink` and `paper` can be combined to recolour a plot.

```{r}
#| output-location: default
#| fig-width: 18
p + theme_minimal(paper = "cornsilk", ink = "navy")
```

---

### Complete themes

`accent` has niche application in `geom_smooth()` and `geom_contour()`.

```{r}
#| output-location: default
#| fig-width: 18
p + theme_classic(accent = "red")
```


---

### Geom elements

`theme(geom)` accommodates the layer settings

```{r}
p + theme(geom = element_geom(
  ink    = "red", 
  accent = "black"
))
```

---

### Geom elements

`ink`/`paper`/`accent` act as described for complete themes

```{r}
#| code-line-numbers: "3-5"
#| collapse: true
element_geom()
```

---

### Geom elements

For lines, we distinguish borders (of polygons) and 'naked' lines.

```{r}
#| code-line-numbers: "6-9"
#| collapse: true
element_geom()
```

---

### Geom elements

Illustration of polygon border versus naked line

```{r}
#| code-line-numbers: "9-16"
binwidth <- 2
ggplot(faithful, aes(waiting)) +
  geom_histogram(
    colour = "black", binwidth = binwidth
  ) +
  geom_density(
    aes(y = after_stat(count * binwidth))
  ) +
  theme(geom = element_geom(
    # Affects histogram
    borderwidth = 0.3,
    bordertype  = "solid",
    # Affects density line
    linewidth = 1,
    linetype  = "dotted"
  ))
```

---

### Geom elements

For text, we can set `family` and `fontsize`

```{r}
#| code-line-numbers: "10-11"
#| collapse: true
element_geom()
```

---

### Geom elements

For text, we can set `family` and `fontsize`

```{r}
#| code-line-numbers: "5-8"
ggplot(mtcars) +
  aes(disp, mpg, label = rownames(mtcars)) +
  geom_label() +
  theme(
    geom = element_geom(
      family = "Barrio",
      fontsize = 10
    )
  )
```

---

### Geom elements

For points, we can set `pointsize` and `pointshape`

```{r}
#| code-line-numbers: "12-13"
#| collapse: true
element_geom()
```

---

### Geom elements

For points, we can set `pointsize` and `pointshape`

```{r}
p + theme(geom = element_geom(
  pointsize  = 2,
  pointshape = "triangle"
))
```

---

### Geom elements

There is also the familiar `colour` and `fill`.

```{r}
#| code-line-numbers: "14-15"
#| collapse: true
element_geom()
```

---

### Geom elements

`colour` and `fill` are direct, but indiscriminate. Note how `accent` is now ignored.

```{r}
p + theme(geom = element_geom(
  colour = "red", 
  fill   = "cyan",
  accent = "limegreen"
))
```

---

### Geom elements

`colour` and `fill` are designed to tailor for individual geom types.

```{r}
p + theme(
  geom.point  = element_geom(colour = "dodgerblue"),
  geom.smooth = element_geom(colour = "navy")
)
```

---

### Ink & Paper: summary

* Complete themes can be used to coordinate a colour scheme quickly.
    * `theme_complete(ink, paper, accent)`
* Most layer defaults can be set in the theme.
    * `theme(geom = element_geom(...))`
    
## [Exercise 4.1](https://posit-conf-2025.github.io/ggplot2/exercises/4_theme_guide.qmd#exercise-4.1-ink--paper) {.center style="text-align:center;"}

{{< countdown minutes=3 top="0" bottom="" >}}

---

## Default palettes

Several ways to set a plot's colour palette

* Directly add a scale
    * `+ scale_colour_gradientn()`
    * Not set as default scale
* Using esoteric options
    * `options(ggplot2.continuous.colour = scale_colour_gradientn)`
    * Arcane
* Overriding default scale
    * `scale_colour_continuous <- scale_colour_gradientn`
    * Messes with namespace
* NEW: via `theme()`

---

### Controlling palettes

A tour through various options

```{r}
#| fig-keep: last
#| code-line-numbers: "8-9|11-14|16-19|21-22"
my_palette <- hcl.colors(15, "sunset")
custom_scale <- function(...) {
  scale_colour_gradientn(colours = my_palette, ...)
}

p2 <- p + aes(colour = cty)

# 1. Directly set a scale
p2 + scale_colour_gradientn(colors = my_palette)

# 2. Using options
options(ggplot2.continuous.colour = custom_scale)
p2
options(ggplot2.continuous.colour = NULL) # reset

# 3. Redefine scale
scale_colour_continuous <- custom_scale
p2
scale_colour_continuous <- ggplot2::scale_colour_continuous # reset

# 4. Using theme
p2 + theme(palette.colour.continuous = my_palette)
```

---

### Controlling palettes

Regular scales override theme defaults

```{r}
p2 + 
  theme(palette.colour.continuous = my_palette) +
  scale_colour_viridis_c()
```

### Controlling palettes

The default palette does not commit to a full scale.
You can still tweak the options via the default scales (`scale_{aesthetic}_{type}`).

```{r}
p2 + 
  theme(palette.colour.continuous = my_palette) +
  scale_colour_continuous(
    breaks = c(
      ten    = 10,
      twenty = 20,
      thirty = 30
    )
  )
```

---

### Theme palettes

The palette settings have the syntax `palette.{aesthetic}.{type}`, where `type` can be `"continuous"` or `"discrete"`.

```{r}
shapes <- c("triangle filled", "square filled")
p +
  aes(
    fill = cty,
    shape = factor(year)
  ) +
  theme(
    palette.shape.discrete = shapes,
    palette.fill.continuous = "sunset"
  )
  
```

---

### Palette input

The palatable palettes are those that can pass through `scales::as_continuous_pal()` and `scales::as_discrete_pal()` respectively.

```{r}
#| collapse: true
#| output-location: default
library(scales)

# Discrete palettes
pal <- as_discrete_pal(c("foo", "bar", "qux"))
pal(2)
palette_type(pal)
palette_nlevels(pal)

# Colours as continuous palette
pal <- as_continuous_pal(c("magenta", "green"))
pal(c(0, 0.5, 1))
palette_type(pal)
```

---

### Palette input

With the right metadata, we can interchange discrete and continuous palettes.

```{r}
#| collapse: true
pal <- new_discrete_palette(
  pal_manual(c("magenta", "cyan", "green")),
  type = "colour", nlevels = 3
)

is_continuous_pal(pal)

con_pal <- as_continuous_pal(pal)
is_continuous_pal(con_pal)

seq(0, 1, length.out = 9) |>
  con_pal() |>
  show_col()
```

---

### Palette input

Just because we can swap discrete and continuous palettes, doesn't mean we should!

```{r}
p2 + theme(
  # No! What have I done?!
  palette.colour.continuous = pal_brewer("qual")
)
```

---

### Default palettes: summary

* New theme arguments to set default palettes: `palette.{aes}.{type}`.
* Input for discrete palettes `scales::as_discrete_palette()`.
* Input for continuous palettes `scales::as_continuous_palette()`.

## [Exercise 4.2](https://posit-conf-2025.github.io/ggplot2/exercises/4_theme_guide.qmd#exercise-4.2-default-palettes) {.center style="text-align:center;"}

{{< countdown minutes=3 top="0" bottom="" >}}

---

## Build your own theme

How do you intend to use a personal theme?

* Complete theme
* Partial theme

---

### Building a complete theme

Capture your theme as a function.

```{r}
my_theme <- function(...) {
  NULL
}

p + my_theme()
```

---

### Building a complete theme

Start with using a complete theme as base.

```{r}
my_theme <- function(...) {
  theme_gray(...)
}

p + my_theme()
```

---

### Building a complete theme

Build your own customisations on top.

```{r}
my_theme <- function(...) {
  theme_gray(...) +
    theme(
      axis.line = element_line(),
      strip.background = element_blank()
    )
}

p + my_theme()
```

---

### Building a complete theme

For custom fonts, put in guardrails for their possible absence.

```{r}
my_theme <- function(..., header_family = "Dancing Script") {
  systemfonts::require_font(header_family)
  theme_gray(..., header_family = header_family) +
    theme(
      axis.line = element_line(),
      strip.background = element_blank()
    )
}

p + my_theme()
```

---

### Building a complete theme

You can coordinate the `ink`/`paper`/`accent` settings, along with the `geom` argument.

```{r}
my_theme <- function(
  ...,
  ink = "hotpink",
  accent = "black",
  header_family = "Dancing Script"
) {
  systemfonts::require_font(header_family)
  theme_gray(..., ink = ink, accent = accent, header_family = header_family) +
    theme(
      geom = element_geom(pointshape = "♥", pointsize = 3),
      axis.line = element_line(),
      strip.background = element_blank()
    )
}

p + my_theme()
```

---

### Building a complete theme

If you have flamboyant `ink` or `paper` colours, you may also want to direct the `palette.colour.continuous` and `palette.colour.discrete` palettes.

```{r}
my_theme <- function(
  ...,
  ink = "hotpink",
  accent = "black",
  header_family = "Dancing Script"
) {
  systemfonts::require_font(header_family)
  theme_gray(..., ink = ink, accent = accent, header_family = header_family) +
    theme(
      geom = element_geom(pointshape = "♥", pointsize = 3),
      axis.line = element_line(),
      strip.background = element_blank(),
      palette.colour.continuous = c("hotpink", "pink", "white")
    )
}

p2 + my_theme()
```

---

### Building a complete theme

You can activate your theme for all plots in your document by using `set_theme()`.

```{r}
set_theme(my_theme())
# Now all plots will be rendered with your theme
p2
```


---

### Building a complete theme

But for now, we'll reset it to something familiar.

```{r}
#| eval: false
set_theme(theme_gray())
p2
```

```{r}
#| echo: false
theme_set(
  theme_gray() +
    theme(
      panel.background = element_rect(colour = NA, fill = "#F0F1EB")
    )
)
p2
```

---

### Building a partial theme

For individual plots, the same advice holds for building a theme function.

However, you can have partial theme 'shortcuts'.

---

### Building a partial theme

You still call a `theme()` or `theme_sub()` function to initiate a fresh, (incomplete) theme.
No need to start with a complete theme base.

```{r}
horizontal_grid <- function() {
  theme_sub_panel(
    grid.major.x = element_blank(),
    grid.minor.x = element_blank(),
    grid.major.y = element_line(),
    grid.minor.y = element_line()
  )
}

p + horizontal_grid()
```

---

### Building your own theme: summary

* Implement as function.
* Use complete theme as basis
* Add on your tweaks
* (Guard custom fonts)
* (Coordinate colour & palettes)

## [Exercise 4.3](https://posit-conf-2025.github.io/ggplot2/exercises/4_theme_guide.qmd#exercise-4.3-build-your-own-theme) {.center style="text-align:center;"}

{{< countdown minutes=5 top="0" bottom="" >}}

---

## Custom guides

ggplot2 has more guides than colour bars and legends.

* axes
* colour bar
* colour steps
* legend
* binned guide

---

### Base guides

A guide can be specified in `guides()` or in the `scale_*(guide)` argument.

```{r}
p +
  aes(size = cty) +
  scale_size_continuous(guide = "bins") +
  guides(x = guide_axis(minor.ticks = TRUE))
```

---

### Styling guides

Guides can have their own local themes for customisation.

```{r}
p + aes(shape = factor(year), colour = class) +
  guides(
    x = guide_axis(theme = theme_classic(ink = "red")),
    y = guide_axis(theme = theme_minimal(ink = "blue")),
    shape = guide_legend(theme = theme_gray(paper = "cornsilk")),
    colour = guide_legend(theme = theme(legend.text.position = "left"))
  )
```

---

### Guide extensions

Specialised guides for niche applications, like periodical data.

```{r}
library(legendry)

dplyr::filter(txhousing, city == "Houston") |>
  ggplot(aes(date, median, colour = month)) +
  geom_line(linewidth = 2, lineend = "round") +
  scico::scale_colour_scico(
    breaks = 1:12, limits = c(0, 12),
    palette = "romaO",
    guide = "colring"
  ) +
  theme(legend.key.size = unit(5, "mm"))
```

---

### Guide extensions

Guides can be specified flexibly to at least the degree of secondary axes.

```{r}
# Translate Celsius range to Fahrenheit
deg_C <- range(pressure$temperature)
deg_F <- (deg_C * 9 / 5) + 32
# Compute breaks and translate back to Celsius
deg_F <- scales::extended_breaks()(deg_F)
deg_C <- (deg_F - 32) * 5 / 9

fahrenheit_axis <- guide_axis_base(
  key = key_manual(aesthetic = deg_C, label = deg_F)
)

p3 <- ggplot(pressure) +
  aes(temperature, pressure) +
  geom_line() +
  theme(axis.line.x = element_line())

p3 + guides(x.sec = fahrenheit_axis)
```

---

### Guide extensions

In addition, many guides can be composed like stacking the Celsius and Fahrenheit guide.

```{r}
stacked <- compose_stack(
  "axis",
  fahrenheit_axis,
  side.titles = c("°C", "°F")
)

p3 + guides(x = stacked)
```

---

### Guide extensions

legendry has re-imagined some guides as compositions.

```{r}
p3 + aes(colour = temperature) +
  guides(colour = guide_colbar(
    second_guide = fahrenheit_axis,
    suppress_labels = "none"
  ))
```


---

### Guide extensions

Aside from composability, re-imagined guides come with extra features.

```{r}
#| dev: svglite
p3 + aes(colour = temperature) +
  scale_colour_viridis_c(
    limits = c(100, 300),
    guide = "colbar"
  )
```

---

### Guide extensions

Aside from composability, re-imagined guides come with extra features.

```{r}
p3 + aes(colour = temperature) +
  scale_colour_viridis_c(
    limits = c(NA, 300),
    guide = "colbar",
    oob = scales::oob_squish
  )
```

---

### Custom guides: summary

* Guides give your plots meaning
* Guides can have local themes for individually tweaking them
* Guide extensions, such as those in the legendry package, expand your arsenal.

## [Exercise 4.4](https://posit-conf-2025.github.io/ggplot2/exercises/4_theme_guide.qmd#exercise-4.4-custom-guides) {.center style="text-align:center;"}

{{< countdown minutes=3 top="0" bottom="" >}}

## {.center}

Next session: [Plot composition](5_composition.qmd)
