---
title: "Interactive plots"
author: "Teun van den Brand"
knitr:
  opts_chunk:
    dev.args:
      scaling: 2
---

```{r}
#| label: setup
#| echo: false
"padding:20px;background:black;color:white;border-radius:8px;text-align:left;font-size:36px" |>
  ggiraph::girafe_css() |>
  ggiraph::opts_tooltip() |>
  ggiraph::set_girafe_defaults(opts_tooltip = _)
```


## Interactivity

---

### ggplot2's graphics ecosystem

```{mermaid}
%%| echo: false
%%| output-location: default
flowchart TD
  A[ggplot2] --> B[grid]
  B --> C[grDevices]
  C --> D(devices)
  D --> E([file])
```

## gganimate

Renders animations of your plot to GIFs.

```{mermaid}
%%| echo: false
%%| output-location: default
flowchart TD
  A[ggplot2] --> B[grid]
  B --> C[grDevices]
  C --> D(devices)
  D --> E([GIF])
  F[gganimate] <--> A
  F -.-> |"animate()"|D
```

---

### basic plot

```{r}
#| output-location: default
suppressPackageStartupMessages({
  library(ggplot2)
  library(gganimate)
  library(gapminder)
})

gapminder <- gapminder |>
  # inoculate against apostrophe in Cote d'Ivoire
  dplyr::mutate(country = gsub("'", "’", country))

gap2007 <- dplyr::filter(gapminder, year == 2007)

gaps <- ggplot(gapminder) +
  aes(
    gdpPercap, lifeExp, 
    size = pop, 
    group = country, 
    colour = continent
  ) +
  geom_point(alpha = 0.7) +
  scale_x_log10(
    guide = "axis_logticks", 
    labels = scales::label_log(),
    minor_breaks = scales::minor_breaks_log(detail = 5),
    limits = range(gapminder$gdpPercap)
  ) +
  scale_colour_manual(values = continent_colors) +
  scale_size_area(
    max_size = 12, 
    guide = legendry::guide_circles(),
    breaks = c(100, 500, 1000) * 1e6,
    labels = c("100M", "500M", "1B")
  ) +
  labs(dictionary = c(
    gdpPercap = "Gross Domestic Product per capita",
    lifeExp   = "Life Expentancy",
    continent = "Continent",
    pop       = "Population"
  )) +
  theme(
    legend.key.size = unit(12, "pt"),
    legend.spacing = rel(1)
  )
```

---

### basic plot

```{r}
(my_plot <- gaps + gap2007)
```

---

### animating

```{r}
anim <- gaps + 
  transition_time(year)

animate(anim, device = "ragg_png")
```


---

## plotly

```{mermaid}
%%| echo: false
%%| output-location: default
flowchart TD
  A[ggplot2] --> B[grid]
  B --> C[grDevices]
  C --> D(devices)
  D --> E([file])
  A --> |"`plotly::ggplotly()`"|F(plotly.js)
  F --> G["htmlwidgets (R)"]
  G --> H([html])
```

---

### Rendering plotly

Using `plotly::ggplotly()` to render interactive plot.

```{r}
suppressPackageStartupMessages(library(plotly))
ggplotly(my_plot)
```


---

### The `text` plotly aesthetic 

The `tooltip` argument gives the aesthetics that contribute to the hover information.
There is an understanding of the `text` aesthetic (note: not `label`).

```{r}
ggplotly(
  my_plot + aes(text = country),
  tooltip = "text"
)
```


---

### The `frame` plotly aesthetic

To animate a plotly graph, use the `frame` aesthetic.

```{r}
ggplotly(
  gaps + aes(frame = year, text = country),
  tooltip = 'text'
)
```



---

## ggiraph

```{mermaid}
%%| echo: false
%%| output-location: default
flowchart TD
  A[ggplot2] --> B[grid]
  B --> C[grDevices]
  C --> D("ggiraph::dsvg()")
  D --> E([SVG])
  F[ggiraph] <--> A
  F -.-> |"ggirafe()"|D
```

---

### Overview ggiraph

* Parallel functions for interactivity
  * `geom_point_interactive()`
  * `element_text_interactive()`
  * `facet_wrap_interactive()`
  * `guide_legend_interactive()`
  * `scale_colour_manual_interactive()`
* Specialist aesthetics/options
  * Aesthetics `tooltip`, `data_id`, `onclick`
  * `ggirafe(options)`

---

### `tooltip` aesthetic

Swapping out `geom_point()` for `geom_point_interactive()`.

```{r}
#| output-location: slide
#| code-line-numbers: 9-12
#| fig-height: 4
library(ggiraph)

gaps <- ggplot(gap2007) +
  aes(
    gdpPercap, lifeExp, 
    size = pop, 
    colour = continent,
  ) +
  geom_point_interactive(
    aes(tooltip = country),
    alpha = 0.7
  ) +
  scale_x_log10(
    limits = range(gapminder$gdpPercap),
    guide = "axis_logticks", 
    labels = scales::label_log(),
    minor_breaks = scales::minor_breaks_log(detail = 5)
  ) +
  scale_colour_manual(values = continent_colors) +
  scale_size_area(
    max_size = 12, 
    guide = legendry::guide_circles(),
    breaks = c(100, 500, 1000) * 1e6,
    labels = c("100M", "500M", "1B")
  ) +
  labs(dictionary = c(
    gdpPercap = "Gross Domestic Product per capita",
    lifeExp   = "Life Expentancy",
    continent = "Continent",
    pop       = "Population"
  )) +
  theme(
    legend.key.size = unit(12, "pt"),
    legend.spacing = rel(1)
  )

girafe(ggobj = gaps)
```

---

### non-aesthetic `tooltip`

Scales can display tooltips in legends.
N.B.: ensure `breaks` and `tooltip` are parallel!

```{r}
#| output-location: slide
#| fig-height: 4
#| warning: false
#| collapse: true
scale_tooltip <- split(gap2007$country, gap2007$continent) |>
  vapply(paste, character(1), collapse = ", ")
str(scale_tooltip)

new_scale <- scale_colour_manual_interactive(
  values  = continent_colors,
  breaks  = names(scale_tooltip),
  tooltip = scale_tooltip
)

girafe(ggobj = gaps + new_scale)
```

---

### `data_id` aesthetic

Using 1:1 IDs

```{r}
#| output-location: default
#| fig-height: 4
girafe(ggobj = gaps + aes(data_id = country))
```

---

### `data_id` aesthetic

Using 1:many IDs

```{r}
#| output-location: default
#| fig-height: 4
girafe(ggobj = gaps + aes(data_id = continent))
```

---

### `data_id` aesthetic

Using many:1 IDs

```{r}
gappairs <- gap2007 |>
  mutate(
    logGDP = log10(gdpPercap),
    logPop = log10(pop)
  ) |>
  ggplot() +
  geom_point_interactive(
    aes(
      x = .panel_x,
      y = .panel_y,
      data_id = country,
      tooltip = country,
      colour = continent
    )
  ) +
  scale_colour_manual(values = continent_colors) +
  ggforce::facet_matrix(vars(lifeExp, logGDP, logPop))

girafe(ggobj = gappairs)
```

---

### `data_id` aesthetic

Let's make a second plot sharing some countries with the first.

```{r}
countries <- c(
  "Senegal", "Mozambique", "Egypt", "Bolivia", "United States", 
  "Brazil", "Czech Republic", "Romania", "Switzerland", "Cambodia", 
  "Japan", "New Zealand", "India", "Australia"
)

heatgap <- gapminder |>
  filter(country %in% countries) |>
  ggplot() +
  aes(
    year, country,
    fill = lifeExp,
    tooltip = round(lifeExp, digits = 1)
  ) +
  geom_tile_interactive() +
  scale_fill_viridis_c()

girafe(ggobj = heatgap)
```

---

### `data_id` aesthetic

```{r}
#| warning: false
library(patchwork)

combined <- 
  gaps + heatgap & 
  aes(data_id = country) &
  theme(legend.position = "none")

girafe(ggobj = combined)
```

---

### Customising ggiraph: tooltips

```{r}
girafe(
  ggobj = gaps,
  options = list(
    opts_tooltip(
      css = girafe_css("border-radius:16px;padding:8px;border-color:white;border-style:solid"),
      use_fill = TRUE
    )
  )
)
```

---

### Customising ggiraph: hovering

```{r}
girafe(
  ggobj = gaps + aes(data_id = country),
  options = list(
    opts_hover(css = girafe_css("fill:white;")),
    opts_hover_inv(css = girafe_css("opacity:0.1;"))
  )
)
```

---

### Customising ggiraph: linking

```{r}
p <- gaps +
  labs(caption = "Source: https://www.gapminder.org/data/") +
  theme(plot.caption = element_text_interactive(
    colour = "darkblue",
    data_id = "any_unique_id",
    hover_css = "fill:blue;stroke:none;text-decoration:underline;",
    onclick = 'window.open("https://www.gapminder.org/data/")'
  ))

girafe(ggobj = p)
```

---

### Customising ggiraph: zooming

```{r}
girafe(
  ggobj = gaps, 
  options = list(opts_zoom(min = 0.5, max = 5))
)
```


---

### Customising ggiraph: selections

```{r}
girafe(
  ggobj = gappairs, 
  options = list(
    opts_selection(
      type = "multiple", 
      only_shiny = FALSE,
      css = "fill:white;"
    )
  )
)
```

---

## Summary

Each package has strengths and weaknesses.

package   | Fidelity | Animation | Tooltips
----------|----------|-----------|---------
gganimate | Yes      | Yes       | No
plotly    | No       | Yes       | Yes
ggiraph   | Yes      | No        | Yes
