---
title: "Exercises for session 4"
---

```{r}
#| label: setup
library(ggplot2)
# Emoji will not always render flawlessly with default device
knitr::opts_chunk$set(dev = "ragg_png")
```

## Ink & Paper

### Exercise 4.1

We'll work with this plot of the penguins dataset.

```{r}
p <- ggplot(palmerpenguins::penguins) +
  aes(body_mass_g, flipper_length_mm, shape = species) +
  geom_point(na.rm = TRUE) +
  labs(
    x = "Body Mass (g)",
    y = "Flipper Length (mm)",
    shape = "Species"
  )
p
```

Pick any of the pre-existing complete themes and make a 'dark mode' variant
of it.

```{r}
#| eval: false
p + ...
```

<details>
<summary>Hint</summary>

Use the `ink` and `paper` arguments in ---for example--- `theme_gray()`.

<details>
<summary>Solution</summary>

A different theme choice is fine, but the important bit is to set `paper` to a dark colour and `ink` to a light colour.

```{r}
p + theme_gray(ink = "white", paper = "black")
```

</details>

</details>

## Default palettes

### Exercise 4.2

This plot has a shape scale with the default palette.

```{r}
p <- ggplot(mpg, aes(displ, hwy, shape = fl)) +
  geom_point(size = 5) +
  scale_shape_discrete(
    breaks = c(
      regular = "r",
      ethanol = "e",
      diesel  = "d",
      premium = "p",
      CNG     = "c"
    )
  ) +
  labs(
    x = "Engine Displacement",
    y = "Fuel efficiency",
    shape = "Fuel type"
  )
p
```

Make a discrete shape palette using emoji.
You can pick a theme like 'farm animals', 'bugs', 'flowers' or other symbols.
Set it as the default palette in the `theme()`.

```{r}
#| eval: false
my_shape_pal <- ...

p + theme(... = my_shape_pal)
```


<details>
<summary>Hint</summary>

Shapes can use emoji because these count as 'single letter' strings.
The palette can be a character vector that is compatible with `scales::as_discrete_pal()`.

<details>
<summary>Solution</summary>

The important bit is to realise that one should populate the `palette.shape.discrete` argument.

```{r}
my_shape_pal <- c("ðŸ«›", "ðŸ“", "ðŸ¥•", "ðŸ‹", "ðŸ‡")

p + theme(palette.shape.discrete = my_shape_pal)
```

</details>

</details>

## Build your own theme

### Exercise 4.3

We have the following contour plot of eruption measurements taken from the geyser 'Old Faithful'.

```{r}
p <- ggplot(faithful) +
  aes(waiting, eruptions) +
  geom_density2d(bins = 8) +
  geom_point() +
  labs(
    x = "Waiting time (min)",
    y = "Eruption time (min)"
  )
p
```

Pick some topic that has a strong association with a colour scheme.
It can be a forest, the ocean, the desert or something you come up with.
Create your own theme with that topic in mind that:

* Is a function
* Passes arguments to a base theme
* Adjusts layer defaults

```{r}
#| eval: false
my_theme <- ...

p + my_theme()
```

<details>
<summary>Hint</summary>

You may want to pass an `ink`/`paper`/`accent` argument to the base theme.
Layer defaults can be set using `element_geom()`.

<details>
<summary>Solution</summary>

This is more of an exercise without a 'correct' answer, but here is an example using the night sky as topic.

```{r}
my_theme <- function(ink = "cornsilk", paper = "#000033", accent = "gold", ...) {
  theme_gray(ink = ink, paper = paper, accent = accent, ...) +
    theme(
      geom = element_geom(pointshape = "\u2605", pointsize = 3, linewidth = 0.2)
    )
}

p + my_theme()
```

</details>

</details>

## Custom guides

### Exercise 4.4

We have the following unemployment data from the United States.

```{r}
library(legendry)

p <- ggplot(economics, aes(date, unemploy)) +
  geom_area(colour = "black", alpha = 0.3) +
  labs(
    x = "Date",
    y = "Unemployment (x1000)"
  )

p
```

Pick some event or period in the range of the plot above.
Create an axis annotation to indicate that event or period.
You may need to construct the appropriate `key` argument for a custom guide.

```{r}
#| eval: false
my_annotation <- ...

p + guides(x = compose_stack("axis", my_annotation))
```


<details>
<summary>Hint 1</summary>

If you're stuck without inspiration, here are some arbitrary events and periods:

* First release of ggplot2 (2007)
* Margaret Thatcher serves as the UK's prime minister (1979-1990)
* Release of the original Sony Walkman (1979)
* The pop group 'The Spice Girls' (1994-2000)
* Fall of the Berlin Wall (1989)
* Release of the pinnacle of telecommunication, the Nokia 3310 (2000)
* The British sitcom 'Mr. Bean' (1990-1995)
* Thomas Lin Pederson was born (1985)
* David Tennant plays the 10th Doctor in 'Doctor Who' (2005-2010)
* The American sitcom 'Golden Girls' (1985-1992)
* The birth of Hadley Wickham (1979)
* Texas Instruments launches its first graphing calculator (1990)

<details>
<summary>Hint 2</summary>

You can combine `legendry::primitive_bracket()` + `legendry::key_range_manual()` for periods.
You can use `legendry::guide_axis_base()` + `legendry::key_manual()` for point-events.

<details>
<summary>Solution</summary>

For a range of dates:

```{r}
my_annotation <- primitive_bracket(
  key = key_range_manual(
    start = as.Date("1977-01-20"),
    end   = as.Date("1981-01-20"), 
    name  = "Carter administration"
  ),
  bracket = "square"
)

p + guides(x = compose_stack("axis", my_annotation))
```

For an event:

```{r}
my_annotation <- guide_axis_base(
  key = key_manual(
    aesthetic = as.Date("1991-09-27"),
    label = "Release 'Terminator 2'"
  ),
  theme = theme(
    axis.ticks.length = unit(6, "mm"),
    axis.ticks = element_line(arrow = arrow(length = unit(1.5, "mm")))
  )
)

# Swapping the composition is a little bit cheating on my end
p + guides(x = compose_ontop("axis", my_annotation))
```

</details>

</details>

</details>

