---
title: "Exercises for session 6"
---

```{r}
#| label: setup
library(ggplot2)
library(ggiraph)
library(patchwork)
# Emoji will not always render flawlessly with default device
knitr::opts_chunk$set(dev = "ragg_png")
```

### Exercise 6.1

We have two plots. One is a plot of the gapminder data.

```{r}
# Wrangle data to add country codes
gapdata <- gapminder::gapminder_unfiltered |>
  dplyr::filter(year == 2007) |>
  dplyr::left_join(gapminder::country_codes, by = "country")

gapplot <- ggplot(gapdata) +
  aes(
    gdpPercap, lifeExp, size = pop, 
    colour = continent, alpha = I(0.7)
  ) +
  scale_colour_manual(
    guide = "none",
    values = gapminder::continent_colors
  ) +
  scale_x_log10() +
  scale_size_area(max_size = 12, guide = "none")

gapplot + geom_point()
```

The second plot is a world map.

```{r}
world <- rnaturalearth::ne_countries()

worldmap <- ggplot(world) +
  aes(colour = region_un, fill = after_scale(alpha(colour, 0.7))) +
  scale_colour_manual(
    guide = "none",
    values = gapminder::continent_colors
  )

worldmap + geom_sf()
```

Note that the `gapplot` and `worldmap` ggplot objects don't have any layers.
Add interactive layers to these two plots, and coordinate them such that hovering over a country in one plot will highlight that country in the other plot too.
Assume that the `world$iso_a3_eh` and `gapdata$iso_alpha` are compatible identifiers for countries.
Optionally, you may also add a `tooltip` to display the country name (in `world$name` and `gapdata$country`).

```{r}
#| eval: false
i_gapplot <- gapplot + ...

i_worldmap <- worldmap + ...

girafe(ggobj = i_gapplot / i_worldmap)
```

<details>
<summary>Hint</summary>

The `geom_point()` and `geom_sf()` layers have parallel `ggiraph::geom_point_interactive()` and `ggiraph::geom_sf_interactive()` layers for interactive use.

The coordination should take place on the level of the `data_id` aesthetic.

<details>
<summary>Solution</summary>

```{r}
i_gapplot <- gapplot +
  geom_point_interactive(aes(
    data_id = iso_alpha, 
    tooltip = country
  ))

i_worldmap <- worldmap +
  geom_sf_interactive(aes(
    data_id = iso_a3_eh, 
    tooltip = name
  ))

girafe(ggobj = i_gapplot / i_worldmap)
```

</details>

</details>

### Exercise 6.2

We have battleship placements on a grid.

```{r}
battleships <- tibble::tribble(
  ~x,  ~y,
  # Patrol Boat
  "A", 1,
  "B", 1,
  # Destroyer
  "D", 1,
  "D", 2,
  "D", 3,
  # Destroyer
  "F", 3,
  "G", 3,
  "H", 3,
  # Battleship
  "G", 1,
  "H", 1,
  "I", 1,
  "J", 1,
  # Patrol Boat
  "J", 3,
  "J", 4,
  # Battleship
  "A", 3,
  "A", 4,
  "A", 5,
  "A", 6,
  # Patrol Boat
  "A", 8,
  "B", 8,
  # Patrol Boat
  "F", 6,
  "F", 7,
  # Destroyer
  "J", 6,
  "J", 7,
  "J", 8,
  # Carrier
  "E", 10,
  "F", 10,
  "G", 10,
  "H", 10,
  "I", 10,
  "J", 10,
)
```

We visualise these battleships with explosions in a plot.
Ignore the horizontal justification for now, this is in anticipation of `girafe()`'s custom SVG device.

```{r}
grid <- expand.grid(y = 1:10, x = LETTERS[1:10])

p <- ggplot(grid, aes(x, factor(y))) +
  geom_text(
    aes(label = "Hit!"), 
    data = battleships,
    size = 8, hjust = 1
  ) +
  geom_text(
    aes(label = "Miss..."), 
    data = vctrs::vec_set_difference(grid, battleships),
    size = 8, hjust = 1
  ) +
  discrete_scale(
    name = "💣",
    aesthetics = "label", 
    palette = scales::pal_manual(c("💥", "🌊"))
  ) +
  scale_y_discrete(limits = as.character(10:1)) +
  theme(legend.title = element_text(size = 25))

p
```

Can you add interactive tiles that cover up the emoji and disappear when selected?
The selection mechanisms go into `girafe(options)`.

```{r}
#| eval: false
p_i <- p + ...

girafe(
  ggobj = p_i,
  options = list(
    ...
  )
)
```


<details>
<summary>Hint</summary>

You need to set the `data_id` aesthetic in `geom_tile_interactive()` so that every cell has a unique ID.
For activating the selection mechanic you can use `opts_selection(css = ..., type = "multiple", only_shiny = FALSE)`.
To make the cells disappear you have to fill in the correct `css` argument.

<details>
<summary>Solution</summary>

```{r}
p_i <- p + geom_tile_interactive(
  aes(data_id = seq_along(x)),
  # Optional styling
  fill = "#EBEBEB", colour = "white", linewidth = 0.5
)

girafe(
  ggobj = p_i,
  options = list(
    # Hover is optional to make it clearer which cell you're about to hide
    opts_hover(css = girafe_css(css = "fill:#888888;")),
    # Trick is to use 0 opacity for selected cells
    opts_selection(css = "opacity:0;", type = "multiple", only_shiny = FALSE)
  )
)
```

</details>

</details>

Without cheating and looking at static plot or the `battleships` data frame, how many clicks does it take you find:

* 1 Carrier (5 cells)
* 2 Battleships (4 cells)
* 3 Destroyers (3 cells)
* 4 Patrol boats (2 cells)
